---
title: "The Effect of Transmission Type on Automobile Fuel Consumption"
author: "Jon Cusick"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Main Questions:  
1. Is automatic or manual transmission better for mpg?  
2. Quantify the difference between the two types of transmission.


This analysis will make use of the `mtcars` (ADD LINK) dataset, a collection of fuel useage for 32 cars, each of which is a 1973-1974 model. An initial exploratory plot shows a high collection of automatic cars at lower MPG levels while manual vehicles are distributed somewhat more evenly at mid to high levels of MPG.

```{r eda, message = FALSE, out.width = "65%"}
library(tidyverse)
library(broom)
library(knitr)

dat <- mtcars %>% select(mpg, am) %>% 
    mutate(Transmission = ifelse(am == 0, "Automatic", "Manual"))

ggplot(dat, aes(x = mpg, fill = Transmission)) +
    geom_density() +
    scale_fill_brewer(palette = "Dark2") + 
    facet_wrap(~ Transmission) +
    labs(x = "Miles per gallon", y = "Density", 
         title = "Distribution of MPG average for 32 vehicles by Transmission Type")
```


A linear regression model was fit without an intercept to the dataset for better intrepretability of both transmission categories. A summary table of the estimates as well as a residual plot can be seen below.

TODO: Add residual plot along x colored by trans

```{r linear_model}
fit <- lm(mpg ~ Transmission + 0, data = dat)
kable(tidy(fit))

err <- resid(fit)


```