---
title: "Exponential Simulation Exercise"
author: "Jon Cusick"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Overview
This report will investigate the exponential distribution ($\mu = \frac{1}{\lambda}, \sigma^2 = (\frac{1}{\lambda})^2$) and look at its properties both from a theoretical and simulated standpoint. We will use the Law of Large Numbers and the Central Limit Theorem to assume that the simulated mean is representative of the population mean as well as the assumption that distribution of simulated data is approximately normal.

#### Simulations
For this exercise, we will perform 1,000 simulations of 40 exponentials, using $\lambda = 0.2$ throughout. For each simulation, the mean and variance of the 40 exponentials will be recorded, as well as the cumulative mean and variance at that point. This will help to demonstrate how the number of simulations influences our inferred judgement of the population distribution. For further analysis in the next sections, the data are saved into the data frame `sim.results`, which includes five columns (simulation number, mean, cumulative mean, variance, and cumulative variance).

```{r simulation, cache = TRUE}

lambda <- 0.2
n = 1000
sim.results <- data.frame(n = seq(1, n), mean = rep(NA, n), cum.mean = rep(NA, n),
                          variance = rep(NA, n), cum.variance = rep(NA, n))
for (i in 1:n) {
    sim = rexp(n, lambda)
    sim.results$mean[i] <- mean(sim)
    sim.results$variance[i] <- var(sim)
    sim.results$cum.mean[i] <- mean(sim.results$mean[1:i])
    sim.results$cum.variance[i] <- mean(sim.results$variance[1:i])
}
```


#### Sample vs. Theoretical Mean
The theoretical mean of the expontial distribution is 5 when $\lambda = 0.2$. The plot below compares this theoretical value (denoted by the vertical black line) with a histogram of the distribution of averages from the simulation. Due to the the Law of Large Numbers, the simulated mean of `r mean(sim.results$mean)` very closely approximates the theoretical value because of the high number of simulations.

```{r mean histogram, message = FALSE, echo = FALSE, out.width = "55%"}
library(ggplot2)
ggplot(sim.results, aes(mean)) + 
    geom_histogram(binwidth = 0.03, fill = "dodgerblue4") +
    labs(x = "Mean", y = "Count", 
         caption = "Figure 1. Distribution of the mean of 40 exponentials") +
    geom_vline(aes(xintercept = 1 / lambda))
```

It can also be seen that as the number of simulations increases, the simulated mean moves closer and closer to the true distribution mean. The plot below shows the cumulative mean (the mean of all simulation results up to and including the current simulation number) as the number of simulations increased.

```{r mean convergence, echo = FALSE, out.width = "55%"}
ggplot(sim.results, aes(x = n, y = cum.mean)) +
    geom_line(color = "dodgerblue4") +
    labs(x = "Number of simulations", y = "Cumulative mean",
         caption = "Figure 2. Simulated mean convergence") +
    geom_hline(aes(yintercept = 1 / lambda))
```

#### Sample vs. Theoretical Variance
The same principle demonstrated with the mean also holds true for the simulated variance. The distribution of variance of the 40 exponentials within each simulation will very closely approximate the true variance of the exponential distribution itself. In the plot below, the vertical line shows exponential variance ($\lambda = 0.2$) of 25 and it is clear that the simulations provided a good approximation, at `r mean(sim.results$variance)`.

```{r variance histogram, echo = FALSE, out.width = "55%"}
ggplot(sim.results, aes(variance)) +
    geom_histogram(binwidth = 0.25, fill = "dodgerblue4") +
    labs(x = "Variance", y = "Count",
         caption = "Figure 3. Distribution of the variance of 40 exponentials") +
    geom_vline(aes(xintercept = (1 / lambda)^2))
```

Again, likewise to the cumulative simulation mean converging to the theoretical mean, the cumulative variance of the simulations begins to converge to the theoretical variance value as the number of simulations increased.

```{r variance convergence, echo = FALSE, out.width = "55%"}
ggplot(sim.results, aes(x = n, y = cum.variance)) +
    geom_line(color = "dodgerblue4") +
    labs(x = "Number of simulations", y = "Cumulative variance",
         caption = "Figure 4. Simulated variance convergence") +
    geom_hline(aes(yintercept = (1 / lambda^2))) 
```


#### Distribution
A common way to compare if two sets of samples come from the same distribution is to use a quantile-quantile plot. These types of figures plot the $n$-th quantile of two distriubtions, using distribution A for the $x$ value and distribution B for the $y$ value. If the quantiles for each plot truly are drawn from the same distribution, it would be expected that every point in the figure falls on the $y = x$. However, if the two distributions are not identical, but related in a linear way, the points are expected to fall along a diagonal line still. Below, we can plot the simulated quantiles against the theoretical to determine that there is indeed a very close match of the two distributions. The only differences are noticed out in the two extreme ends of the simulated distribution.

```{r distribution q-q, echo = FALSE, out.width = "55%"}
ggplot(sim.results, aes(sample = mean)) +
    stat_qq(color = "dodgerblue4", distribution = qnorm) +
    labs(x = "Theoretical", y ="Sample", 
         caption = "Figure 5. Q-Q plot of exponential simulated means distribution vs. normal distribution")
```



#### Appendix A: Plotting code

Figure 1
```{r a mean histogram, message = FALSE, eval = FALSE, out.width = "60%"}
library(ggplot2)
ggplot(sim.results, aes(mean)) + 
    geom_histogram(binwidth = 0.03, fill = "dodgerblue4") +
    labs(x = "Mean", y = "Count", 
         caption = "Figure 1. Distribution of the mean of 40 exponentials") +
    geom_vline(aes(xintercept = 1 / lambda))
```

Figure 2
```{r a mean convergence, eval = FALSE, out.width = "60%"}
ggplot(sim.results, aes(x = n, y = cum.mean)) +
    geom_line(color = "dodgerblue4") +
    labs(x = "Number of simulations", y = "Cumulative mean",
         caption = "Figure 2. Simulated mean convergence") +
    geom_hline(aes(yintercept = 1 / lambda))
```


Figure 3
```{r a variance histogram, eval = FALSE, out.width = "60%"}
ggplot(sim.results, aes(variance)) +
    geom_histogram(binwidth = 0.25, fill = "dodgerblue4") +
    labs(x = "Variance", y = "Count",
         caption = "Figure 3. Distribution of the variance of 40 exponentials") +
    geom_vline(aes(xintercept = (1 / lambda)^2))
```


Figure 4
```{r a variance convergence, eval = FALSE, out.width = "60%"}
ggplot(sim.results, aes(x = n, y = cum.variance)) +
    geom_line(color = "dodgerblue4") +
    labs(x = "Number of simulations", y = "Cumulative variance",
         caption = "Figure 4. Simulated variance convergence") +
    geom_hline(aes(yintercept = (1 / lambda^2))) 
```

Figure 5
```{r a distribution q-q, eval = FALSE, out.width = "60%"}
ggplot(sim.results, aes(sample = mean, color = "dogerblue4")) +
    stat_qq(color = "dodgerblue4", distribution = qnorm) +
    labs(x = "Theoretical", y ="Sample", 
         caption = paste0("Figure 5. Q-Q plot of exponential simulated means ",
                            "distribution vs. normal distribution"))
```

